A dynamic bioenergetic model for coral-*Symbiodinium* symbioses and coral bleaching as an alternate stable state

### 2017-02-07 JTB_ms/JTB_ms.Rmd
fix footnote duplication

## Seasonal variability
Symbiont densities and coral growth rates are known to vary seasonally, representing an integrated response to changes in a suite of environmental factors. Light in particular is a strong driver of these trends [@Stimson:1997p3837; @Brown:1999p3534; @Fagoonee:1999p4136; @Fitt:2000p3792], with high light associated with lower symbiont abundance and reduced tissue biomass. The role of light in driving seasonal changes in symbiont density was demonstrated nicely by @Stimson:1997p3837, who also found that experimental nutrient-enrichment amplified the light-driven seasonal oscillation. Using the levels of light and nutrients from this study as inputs, the model reproduces this observed interaction among environmental factors (Fig. 5), and also provides the mechanism: as increasing light in summer decreases carbon fixation due to photooxidative stress, symbionts grow slower than the host, leading to decreasing S:H ratios. Under nutrient enrichment, this effect is more pronounced, as CO~2~-limitation (due to higher symbiont standing stocks) causes mild bleaching that results in a similar summertime minimum S:H as the ambient nutrient case ('physiological bleaching' *sensu* @Fitt:2001p832). Decreasing light into winter then alleviates the photooxidative stress constraints on carbon fixation, such that nitrogen constrains the S:H ratio, explaining why S:H increases more when DIN is enriched (Fig. S8). The model of @Muller would also predict higher S:H ratios in winter if light became limiting to photosynthesis, but this is distinct from the mechanism here, in which stress in the summer drives a reduction in S:H. These can be distinguished as follows:

winter increase in S:H as light limits carbon fixation - decreases growth in winter
summer decrease in S:H as photo-oxidative stress limits carbon fixation - decreases growth in summer

under high light in summer, photooxidative stress decreases symbiont biomass and increases carbon-limitation of host growth regardless of nutrient status; when light is reduced in winter, growth and symbiont biomass become constrained instead by nitrogen, and therefore increase more when DIN is enriched (Fig. S8).

@@

The prediction of higher growth when light is reduced indicates that growth is not limited by low light in winter, but is actually reduced by excess light in summer[^2], consistent with the experimental findings of @Schutter:2011ie. Seasonal summertime reductions in tissue biomass have also been well-documented in the field [@Fitt:2000p3792], along with reductions in net photosynthetic capacity [@MullerParker:1987cr]. Importantly, while light alone may drive seasonal dynamics in the ways discussed[^2], temperature fluctuations may attenuate or even reverse the effect of light as cooler winters depress metabolism; thus, the relative magnitude of fluctuation in temperature, light, and other factors may produce wide variability in the direction and magnitude of seasonal changes in growth and symbiont abundance, depending on location and microhabitat. Nevertheless, the seasonal variability predicted here (Fig. 5) is consistent with experimental and field observations for corals, and demonstrates the power of this model to predict dynamic behavior that mechanistically integrates multiple environmental drivers.

[^2]: at the light levels indicated in @Stimson:1997p3837. Note that if light levels were reduced throughout the year (e.g. for a coral at greater depth) such that light did not cause photo-oxidative stress in summer but became limiting to photosynthesis in winter, the S:H ratio would still increase in winter, but growth would decrease; the latter scenario is predicted both by the present model as well as that of @Muller:2009io, since photo-oxidative stress does not become relevant.

>>>>

## Seasonal variability
Symbiont densities and coral growth rates are known to vary seasonally, representing an integrated response to changes in a suite of environmental factors. Light in particular is a strong driver of these trends [@Stimson:1997p3837; @Brown:1999p3534; @Fagoonee:1999p4136; @Fitt:2000p3792], with high light associated with lower symbiont abundance and reduced tissue biomass. The role of light in driving seasonal changes in symbiont density was demonstrated nicely by @Stimson:1997p3837, who also found that experimental nutrient-enrichment amplified the light-driven seasonal oscillation. Using the levels of light and nutrients from this study as inputs, the model reproduces this observed interaction among environmental factors (Fig. 5), and also provides the mechanism: as increasing light in summer decreases carbon fixation due to photooxidative stress, symbionts grow slower than the host, leading to decreasing S:H ratios. Under nutrient enrichment, this effect is more pronounced, as CO~2~-limitation (due to higher symbiont standing stocks) causes mild bleaching that results in a similar summertime minimum S:H as the ambient nutrient case ('physiological bleaching' *sensu* @Fitt:2001p832). Decreasing light into winter then alleviates the photooxidative stress constraints on carbon fixation, such that nitrogen constrains the S:H ratio, explaining why S:H increases more when DIN is enriched (Fig. S8).

@@

The prediction of higher growth when light is reduced indicates that growth is not limited by low light in winter, but is actually reduced by excess light in summer[^3], consistent with the experimental findings of @Schutter:2011ie. Seasonal summertime reductions in tissue biomass have also been well-documented in the field [@Fitt:2000p3792], along with reductions in net photosynthetic capacity [@MullerParker:1987cr]. Importantly, while light alone may drive seasonal dynamics in the ways discussed[^3], temperature fluctuations may attenuate or even reverse the effect of light as cooler winters depress metabolism; thus, the relative magnitude of fluctuation in temperature, light, and other factors may produce wide variability in the direction and magnitude of seasonal changes in growth and symbiont abundance, depending on location and microhabitat. Nevertheless, the seasonal variability predicted here (Fig. 5) is consistent with experimental and field observations for corals, and demonstrates the power of this model to predict dynamic behavior that mechanistically integrates multiple environmental drivers.

[^3]: at the light levels indicated in @Stimson:1997p3837. Note that if light levels were reduced throughout the year (e.g. for a coral at greater depth) such that light did not cause photo-oxidative stress in summer but became limiting to photosynthesis in winter, the S:H ratio would still increase in winter, but growth would decrease; the latter scenario is predicted both by the present model as well as that of @Muller:2009io, since photo-oxidative stress does not become relevant.
